---
title: 'Computer Workshop 7: Linear models with a multiple predictors'
author: "Roos & Pinard"
date: 'Last updated: `r Sys.Date()`'
output:
  word_document:
    toc: yes
    reference_docx: "../sbstemplate.docx"
---

# Workshop Structure

As always, we’re using our tried-and-tested workflow to guide today’s workshop:

1. Formulate a research question.
2. Perform exploratory data analysis (EDA).
3. Identify any hidden assumptions.
4. Fit an appropriate model [today’s main focus].
5. Diagnose the model and check assumptions.
6. Summarise the results.
7. Interpret and draw conclusions.

```{r, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, results = "show", fig.show="show")
answer_sheet <- TRUE
library(ggplot2)
library(here)
```

### A Slight Tweak to Our Workflow

Today, we're going to explore two different approaches to modelling using the same dataset with both models being fit using a linear model. You might recognize this dataset from the first week of the course --- it contains information about rental Prices in Aberdeen.

The two models we're focusing on are:

1.  **A predictive model** aimed at predicting rent Prices as accurately as possible.
2.  **A causal model** designed to understand the effect of square meters on rent Price.

This exercise will show that the same statistical tools can be used for different goals --- whether you want to build a simple predictive "AI" model or explore causal relationships.

# Predictive Modelling

Predictive models are a great starting point when considering models with multiple explanatory variables because they're usually more straightforward --- no need to worry about tricky causal assumptions here (don't tell the Machine Learning folks I said that!).

```{r, echo = FALSE}
abdn <- read.table(here("Workshops/Workshop 2 - Data vis/Data", "abdn_flats.txt"), header = TRUE)
```

## 1. Formulate a Research Question

This one's simple: We just want to predict rent Price. We're not interested in why a rented property (either a flat or house) is more or less expensive; we just want to predict the listed Price as accurately as possible. To do so we will limit ourselves to using the floor area (m$^2$), total number of rooms, the energy efficiency and the council tax band as explanatory variables in the model.

## 2. Perform Exploratory Data Analysis (EDA)

As usual, spend a few minutes exploring the dataset. Check for any obvious errors or odd values. Since you have a link to each property, you can even verify unusual entries yourself; a rare luxury in science!

As a warning, there are properties that when listed were given a specific Price but by the time you look at a particular link, it may have had its Priced reduced. For example, there is one property that when I first added it to the dataset was listed at a whopping £3,200 a month but at the time of writing this, that Price had dropped to £2,900 (because no one was willing to pay £3.2k in rent). In these cases, you'll need to decide for yourself how you want to deal with these. Your options are; 1) retain the original Price if you think it wasn't a data entry mistake, 2) update the Price with the newly lowered rent, or 3) create an entirely new observation for this "new" listing. Note that option 3) will lead to an assumption being violated. Do you know which assumption would be broken if creating a new observation for any properties that had their Price lowered?

## 3. Identify Any Hidden Assumptions

Even in a predictive model, we need to think about potential hidden assumptions. If our data is flawed or not representative, our predictions will be flawed too. Take a moment to consider limitations. For example, can we use this model to predict rent Prices in London?

Another hidden aspect of the data that you may want to consider (and potentially resolve if you are unhappy with it), is that the model will ignore that rental prices are included for both houses and flats, with no distinction made between them. We could always add this in (given we have the data) but let's pretend we didn't have this information. What do you think this lack of distinction will do in our predictive model?

```{r, include=answer_sheet, class.source="bg-success", eval = F}
# A property being either an entire house or just a flat will almost certainly have a big impact on how much the landlord asks for in rent. By not including this in the model, we're leaving an "important" variable out of the model, so our predictions are likely going to be less accurate. Maybe by a lot, maybe by a little, but definitely by "some" amount.

# The other issue is that this will likely create some unequal variance of errors, with predictions being worse (i.e. less accurate, i.e. more residual error) at higher predicted rent Prices (as most houses cost more to rent, the model will likely struggle to understand *why* they are so expensive [or cheap], leading to more error).
```

## 4. Fit an Appropriate Model

One reason predictive modelling is so accessible is that we can throw all of our (sensible) explanatory variables into the model without worrying too much about the underlying relationships (obviously, we wouldn't include the URL as an explanatory variable in the model). Here are all of the variables we have access to:

1.  **`Price`** - the rent of the listed property (£)

2.  **`Bedrooms`** - the number of bedrooms in the property

3.  **`PublicRooms`** - the number of living rooms in the property

4.  **`Rooms`** - the total number of bedrooms plus living rooms in the property

5.  **`Bathrooms`** - the number of bathrooms

6.  **`FloorArea`** - the total floor area of the property (measured in m$^2$)

7.  **`EPC`** - the Energy Performance Certificate (A is highly energy efficient, G is terrible)

8.  **`Tax`** - the council tax band (which is based on the value of the property on 1/4/1991, A is the lowest council tax, H is the highest possible)

9.  **`DayAdded`** - the day when the record was *added to the dataset* (note this is not when the property was listed on the site)

10. **`PropertyURL`** - web link for the property listing

11. **`Address`** - the property name or address

12. **`Latitude`** - the latitude coordinates of the property (note latitude is North to South - think of a "ladder", sounds kinda like "latter" going up)

13. **`Longitude`** - the longitude coordinates of the property (note longitude is East to West - think of "the *long* way around")

14. **`UniCommuteTime`** - the estimated time to walk to the Zoology Building (extracted using the Google Maps API)

15. **`HouseType`** - whether the property is "[detached](https://en.wikipedia.org/wiki/Single-family_detached_home)", "[semi-detached](https://en.wikipedia.org/wiki/Semi-detached)", "[terrace](https://en.wikipedia.org/wiki/Terraced_house)", or "[flat](https://en.wikipedia.org/wiki/Apartment)" (note that "flats" is what British people call "apartments")

16. **`Furnished`** - whether the property is fully furnished, part furnished, or unfurnished

But let's keep it simple today. As noted above, we'll only include the following explanatory variables in the model:

-   `FloorArea`
-   `Rooms`
-   `EPC`
-   `Tax`

Something to note in this dataset are the `EPC` and `Tax` variables. In both of these variables, the actual underlying "values" are continuous (tax band is a measure of the property value in 1991, and EPC is a measure of how energy efficient a property is) but these have been discretised, or converted into groups (e.g. a property valued at £100,000 in 1991 is in the `A` tax band).

> For some reason, the value of a property in 1991 is how the council tax is decided. I don't get why 1991 is special (I doubt it actually is) but that's what they do. It's dumb. C'est la vie.

Whenever this discretisation is done, when it comes to the actual model, it can make our life pretty horrible, which you can see when we write out the full notation of our model. In our case, here's what it looks like:

$Price_i \sim Normal(\mu_i, \sigma)$

$\mu_i = \beta_0 + \beta_1 \times FloorArea_i + \beta_2 \times Rooms_i +$

$\beta_3 \times EPCB_i + \beta_4 \times EPCC_i + \beta_5 \times EPCD_i +$

$\beta_6 \times EPCE_i + \beta_7 \times EPCF_i + \beta_8 \times EPCG_i +$

$\beta_9 \times TaxB_i + \beta_10 \times TaxC_i + \beta_{11} \times TaxD_i +$

$\beta_{12} \times TaxE_i + \beta_{13} \times TaxF_i + \beta_{14} \times TaxG_i$

Where:

-   $\beta_0$ is the intercept, representing a property with zero square meters, zero rooms, an EPC of A, and a Tax band of A.
-   $\beta_1$ is the slope for each additional square meter in floor area.
-   $\beta_2$ is the slope for each additional room.
-   $\beta_{3,...,8}$ are the effects of each EPC rating compared to A.
-   $\beta_{9,...,14}$ are the effects of each Tax band compared to A.

As you can see, `EPC` and `Tax` being recorded as categories has made our model stupidly and crazily more complicated than it would otherwise be. If EPC and Tax were continuous variables, we'd need to estimate just *four* parameters instead of 14! In this case, we can't go backwards. The data we have is categorical and we have to respect that, so we're stuck having to estimate 14 parameters.

A general tip: avoid creating categorical variables out of continuous ones when possible --- it'll save you headaches down the road (and often makes your analyses more reliable).

Despite the complexity, running this model in R is straightforward:

```{r}
rent_mod <- lm(Price ~ FloorArea + Rooms + EPC + Tax, data = abdn)
```

Now let's take a look at the estimated parameters using `summary()`:

```{r}
summary(rent_mod)
```

This produces a giant list of estimated parameters; reminding us why it's best to avoid excessive categorization. Keep this in mind for your future work (including your Honour's project), both for simplicity and for more meaningful analyses!

The complication with predictive models is that these parameters estimates don't have a "biological" or "real world" meaning. We wouldn't read these as "increasing X causes Y to increase" by whatever the parameter value is. Instead it's what the model would predict the increase would be. If that's unclear the lecture slides on causal inference may help clarify.

```{r, echo = F}
beta_mod <- unname(coef(rent_mod)["FloorArea"])

gbp <- scales::label_dollar(prefix = "£", big.mark = ",", accuracy = 0.01)
```

## 5. Diagnose the model and check assumptions.

In previous workshops, we've saved assumption checks for the Friday session, but by now, you should be feeling more confident and able to do this more quickly.

Let's run a quick diagnostic check:

```{r}
par(mfrow = c(2,2))
plot(rent_mod)
par(mfrow = c(1,1))
```

### Question Set One

1.  Do you see any assumptions that are likely broken in the Residuals vs Fitted and Scale-Location plots?

```{r, include=answer_sheet, class.source="bg-success", eval = F}
# Yes, it looks like we have unequal variance of errors in both figures.
```

2.  If so, what is the consequence of violating this assumptions?

```{r, include=answer_sheet, class.source="bg-success", eval = F}
# Our confidence intervals will either be artificially inflated or deflated (i.e. either too small or too big), making our predictions look more or less uncertain than they should be.
```

3.  Do you see any assumptions that are likely broken in the Q-Q Residuals plot?

```{r, include=answer_sheet, class.source="bg-success", eval = F}
# Our residual errors are not perfectly Normally distributed, but they’re not that bad. We have three rental properties that are either more expensive or cheaper than we’d expect given the information provided to the model. If it were much more severe (e.g., dots all over the place), it could lead to inflated 95% confidence intervals.
```

4.  Are there any observations that are having a disproportionate affect on the model?

```{r, include=answer_sheet, class.source="bg-success", eval = F}
# Yes, a few. We can see these in the Residuals vs. Leverage plot, with a few points highlighted by the text output (note that these don’t appear on the plot but are highlighted by the text output).
```

5.  If there are any such observations, what should we do about them (if in doubt, ask a member of staff)?

```{r, include=answer_sheet, class.source="bg-success", eval = F}
# We have three options, though one isn’t very practical for us today:

# 1. Collect more data: This is the ideal solution. Influential observations are usually more problematic in smaller datasets, so gathering more data could help smooth out their impact. Unfortunately, that’s not an option for us today.

# 2. Remove these observations and rerun the analysis: A lot of papers go with this approach. However, the problem comes when they don't make it clear that they've done this, or they remove the observations but don't show the what difference it made. This is not a good option. If you are going to remove the observations, then instead do option 3.

# 3. Rerun the model without these problematic observations and compare the parameter estimates of the two models. If removing them doesn’t drastically change the estimates, you might as well stick with the original model (telling your reader that you've checked both models and showing them both). If it does make a difference, things get more complicated. In that case, you should still report both models but focus on the one you feel is “better.” What “better” means here is up to you; there’s no definitively "right" answer (but not having high leverage might be a sensible measure).

# Here’s how you’d do option 3:
abdn_trimmed <- abdn[-c(324, 534),]
trimmed_rent_mod <- lm(Price ~ FloorArea + Rooms + EPC + Tax, data = abdn_trimmed)
coef(trimmed_rent_mod) # coef() gives just the "coefficients" or parameter estimates
coef(rent_mod)
# By comparing the two models, you can see the effect these observations had. Sometimes this will have an impact, in which case it might seem that there are influential observations but other times the parameter estimates won't really change. Over the years of adding observations to this dataset, sometimes there are flats that once removed changes things but with an increasingly large dataset, they become less impactful.
```

Given your dataset may be slightly different to the one I used above (given how you dealt with any changes in listed Price, houses versus flats, etc.), you may have seen some issues in the Residuals verses Leverage plot (though you may not, which would be fine). If you did see any problematic observations, before moving ahead, remove any observations from your dataset where the associated Cook's distance was $\geq 1$ and rerun your model. As a hint, if I believed the flats on rows 1 and 2 of the dataset were causing issues in the model, I could remove them by doing (observations 1 and 2 chosen here at random just to demonstrate how to remove them):

```{r, eval = F}
abdn <- abdn[-c(1, 2),]
```

The `-` before `c()` basically means "remove", so here we're removing rows 1 and 2 (remember that square brackets are organised as `[row,column]`).

Having removed these and rerun the model (unless your previous checks suggested no issues and you did not need to remove any observations). Do a quick check of the diagnostic plots to see if this has resolved the issue.

```{r, echo = FALSE}
abdn_trimmed <- abdn[-c(324, 329),]
rent_mod <- lm(Price ~ FloorArea + Rooms + EPC + Tax, data = abdn_trimmed)
plot(rent_mod)
```

Hopefully removing those observations has resolved the disproportionately influential observations. There's some important nuance about removing observations from a dataset. For details on this, see the answer sheet for question 5 above.

## 6. Summarise the results.

As usual, pull up the summary table:

```{r}
summary(rent_mod)
```

Pay attention to the `Estimate` for `FloorArea`; the model seems to be predicting that adding an extra 1 square meter to a property only increases the rent by about `r gbp(beta_mod)`. Seems a bit low, right? Do we really think that the difference in Price between a 50 m$^2$ flat and a 100 m$^2$ flat will be just `r gbp(beta_mod * 50)` extra per month? We'll come back to this after we run the causal version of this model.

Given the number of explanatory variables we have in our predictive model, we'll use `ggeffects` to do the plotting, though you can definitely do this manually if you prefer more control (see the instructions with answers to see how to do that).

```{r}
library(ggeffects)
```

Then ask `ggeffects` to show the predicted relationship (or association) between rent Price and each of our explanatory variables:

```{r}
preds <- ggpredict(rent_mod)
```

Now we just need to plot it (note you may need to press the blue arrow buttons in the Plots pane in RStudio to page through the different figures):

```{r}
plot(preds)
```

We get the same message that we saw previously, where `ggeffects` isn't terribly happy using strings that are recorded as `characters` rather than `factors`. If you want to appease the `ggeffects` gods, feel free to convert your categorical variables to `factors` (using the `factor()` function) and rerun your analysis.

```{r, include=answer_sheet, class.source="bg-success", eval = F}
# To make these predictions manually we still do things the same way as before. The only extra complication is that we do this for each explanatory variable in turn. Additionally, we need to specify values for all explanatory variables, even those we aren't making predictions for:

fake <- data.frame(
  FloorArea = seq(from = min(abdn$FloorArea), to = max(abdn$FloorArea), by = 5),
  Rooms = median(abdn$Rooms),
  Tax = "D",
  EPC = "D"
)

preds <- predict(rent_mod, newdata = fake, se.fit = TRUE)
fake$fit <- preds$fit
fake$low <- preds$fit - preds$se.fit * 1.96
fake$upp <- preds$fit + preds$se.fit * 1.96

# Plot
ggplot(fake) +
  geom_ribbon(aes(x = FloorArea, ymin = low, ymax = upp), fill = "lightgrey", alpha = 0.8) +
  geom_line(aes(x = FloorArea, y = fit)) +
  labs(x = "Square meters",
       y = "Predicted rental Price") +
  theme_minimal()
```

### Question Set Two

1.  Can we identify any causal relationships from this model (e.g. "If you increase the square meters of a property by 10 meters squared, it will *cause* the rent to increase by X")?

```{r, include=answer_sheet, class.source="bg-success", eval = F}
# No, we can't identify causal relationships with this model. This model is designed to predict rent Prices and not to understand the cause-and-effect relationship between variables.
```

2.  Why might the relationship between `Price` and `FloorArea` be smaller than we'd expect?

```{r, include=answer_sheet, class.source="bg-success", eval = F}
```

3.  Go to the letting agency website, [ASPC](https://www.aspc.co.uk/), and find a recently listed rental property which has all of the explanatory variables you need. Using your selected property, see how well your model predicts the rent. You're also welcome to use your own flat if you know (or can guess) how many square meters it has, what the EPC rating is, and which Tax band it falls in.

```{r, include=answer_sheet, class.source="bg-success", eval = F}
#E.g. I'll use this property: https://www.aspc.co.uk/search/property/427426/6-Joss-Court/Aberdeen/

# I need to create a dataframe to contain the relevant information:
joss_court <- data.frame(
  Rooms = 3,
  FloorArea = 61,
  EPC = "C",
  Tax = "C"
)

predict(rent_mod, newdata = joss_court)

# The model predicts this property should be £740, and is actually listed at £800; £60 off is not bad for predicting the rent of a flat the model has never seen before! (That £60 is also the residual error!)

# We could always include 95% CI to see if the true rent is within our intervals (using a shortcut to getting 95% CI):
predict(rent_mod, newdata = joss_court, interval = "confidence")
# And the true rent is comfortably within our intervals
```

If you want to see how this kind of model could be translated into something you might see online with some marketing saying something like "We're using the most advanced Generative Artificial Intelligence to predict how much rent you should be asking for", see [here](https://deonroos.shinyapps.io/BI3010_Rent/). This is a little toy website I made (entirely in `R`). There are two important things to note; 1) I think there's a decent chance the website will crash if lots of people try to use it at the same time, and 2) I include another model that deals with space - i.e. certain areas in Aberdeen are more expensive than others. One assumption we're likely to break here is the assumption of independent errors, and including space in the model helps to account for that (this is the "Fancy model" on the site).

# Causal Modelling

Having run a model that aims to predict rent Price, we're now going to switch over to causal modelling using the same dataset. Hopefully the differences and similarities between the two approaches will become clear.

To be responsible, be sure to re-load the data again, so that any changes we made to it specifically for the predictive modelling are not carried into our new analysis. However, if there were any issues that you previously resolved, do so again now.

```{r, echo = FALSE}
abdn <- read.table(here("Workshops/Workshop 2 - Data vis/Data", "abdn_flats.txt"), header = TRUE)
```

## 1. Formulate a Research Question

Our research question is going to be to understand the causal effect of square meters on rent Price, or "How much would rent be if a property were larger or smaller?". The important thing here is that we're going to try and and ask a *causal* question and not simply exploit associations (i.e. correlations) to make accurate predictions.

Because we've already carried out an EDA and thought about hidden assumptions with regards to this data, we can skip these steps but before we get to fitting the model, we have an extra step when we want a model to potentially understand causation.

## Creating a DAG

Because our question is causal, we can't just use the approach we did above and throw everything into the model and read off every parameter estimate as being causal (a practice that is called the "Table 2 Fallacy", so named because the second table of a paper often contains parameter estimates, which authors may then read as being causal). Instead, we need to think about confounding and how to responsibly deal with that in our model. One of the least technical ways of doing this (but not the only way) is to create a Directed Acyclic Graph (or DAG) and use this in combination with the Backdoor Criterion to inform which explanatory variables should, could or should not be included in the model to understand the causal effect of interest. This process of drawing a DAG can sometimes be intellectually challenging (e.g. does the square meters of a property cause the number of rooms, or is it the number of rooms that cause the square meters?) and is often be a pretty laborious process. The intellectual challenge in drawing a DAG is one of the big downsides (and a good argument to use experiments when possible when trying to determine causation), but just drawing the DAGS can be a real pain especially if you create your DAG in PowerPoint. Thankfully there's a much easier way to do this in `R` using the package we've used previously - `performance`.

However, a slight complication is that we're using a "developer" version of `performance` that hasn't yet been put onto CRAN (the central repository where `R` packages are kept), so we'll need to install the package directly from the developers. To do so, we'll need to do a minor tweak to our `install.packages()` code where we say where the version is available from. At the same time, we'll need to download an extra package called `see` from the same developers, which works with `performance` on some things in the background.

To install the correct version of `performance` re-install it by running:

```{r, eval = F}
install.packages(c("performance", "see"), repos = "https://easystats.r-universe.dev")
```

In R, the function `check_dag()` from the `performance` and `see` packages helps us create and evaluate a DAG. Here's how it works:

1.  Specify the Causal Relationships for the Response Variable:

-   You need to outline what you believe the causal relationships are. For example, if you think rent is influenced by the number of rooms, square meters, Tax band, and EPC, you would "translate" this to `Rent ~ Rooms + FloorArea + Tax + EPC`. Note that the variable names in the model don't have to match those in your dataset; just avoid using spaces in variable names, e.g., use `Aberdeen_Rent` instead of `Aberdeen Rent`.

2.  Define the Relationships Between Variables:

-   Having done that, we now just go through what might cause our *explanatory* variables, for example the number of rooms you can have in a flat is must surely be caused by how many square meters there are, so we include this as: `Rooms ~ FloorArea`, and so on for the other explanatory variables.

3.  Set Up the Response and Explanatory Variables:

-   The final part we need to complete is to define what our response variable is (called "`outcome`" by performance) and what our explanatory variable is (called "`exposure`"). This defines our initial and simple model, which in this case would be: `Rent ~ FloorArea`.

Here's what your code might look like (but note that your DAG might look different from mine, e.g. maybe you don't agree with me that EPC is caused by the size of a property):

```{r, message = F}
library(performance)
library(see)
rent_dag <- check_dag(
  Price ~ Rooms + FloorArea + Tax + EPC,
  Rooms ~ FloorArea,
  EPC ~ FloorArea,
  Tax ~ FloorArea,
  exposure = "FloorArea",
  outcome = "Price"
)
```

Once we're happy that we have a DAG that seems sensible, we can then plot it:

```{r, message = FALSE, warning = FALSE}
plot(rent_dag)
```

> Note that the exact positions of the variables in the figure are semi-random, so your figure might not look exactly the same as mine, e.g. `FloorArea` might be on the right instead of the left of `Tax` but the arrow should still go from `FloorArea` to `Tax`.

The left hand figure shows the version of the DAG if we ran the model:

```{r, eval = F}
lm(Price ~ FloorArea, data = abdn)
```

while the figure on the right shows the version of the DAG that would required if we wanted to understand the *total causal effect* that changing `FloorArea` would have on the `Price`. In this case, they look identical? Why is that? Well, note that `FloorArea` drives a lot of causal relationships in our DAG: `FloorArea` → `Rooms` → `Price`, and `FloorArea` → `Tax` → `Price`, etc. So according to this DAG, if a landlord increased the square meters of their property it will have lots of "knock on" effects, e.g. increasing `FloorArea` will increase how many rooms you have, resulting in a higher listing Price. These are all "piped confounds" (sometimes called "mediators"). Remember the example of what to do with pipes? We normally want information to "flow", so we don't "block" them. Here we keep them "unblocked" by not including the respective explanatory variable.

The reason for doing so here is that all of these ways in which changing `FloorArea` will *cause* a change in `Price` are relevant to understanding the *total causal effects* of changing `FloorArea`. If we've increased the square meters of a property, and that results in an extra room, isn't that part of how `FloorArea` causes a `Price` change? So in this case, the model we want to run, to identify the *total causal effect* that changing `FloorArea` would have on the `Price` is:

```{r, eval = F}
lm(Price ~ FloorArea, data = abdn)
```

> Note that it won't always work out that to understand the causal effect of X on Y requires the model Y \~ X. In fact, most of the time it doesn't. It's just that according to this particular DAG, to understand FloorArea on Price we would run the model Price \~ FloorArea.

For a slightly more detailed summary of the DAG, we can simply pull up the DAG object in `R` to use it to identify which model we should run to answer our research question:

```{r}
rent_dag
```

We get two pieces of information here

This tells us that to estimate for the *direct* effect of `FloorArea` on `Price`, we would also need to include EPC and Rooms, but we aren't interested in just the direct effect. For clarity, the *direct* effect would be the effect of adding extra square meters to a property not including the fact that this will likely influence the number of rooms, the EPC, the Tax band, etc. Instead, based on our research question we want to understand what would happen to `Price` if we increased a property by, say, 10 square meters, and all the different ways those extra 10 m result in a change in Price. So we want the *total* effects. For that, `check_dag()` has suggested that the initial model is correctly specified, where our initial model was `Price ~ FloorArea`.

## 4. Fit an appropriate model.

Let's run that model:

```{r}
rent_causal_mod <- lm(Price ~ FloorArea, data = abdn)
```

## 5. Diagnose the model and check assumptions.

And we can do a quick check of the assumptions related to the model residuals:

```{r}
par(mfrow = c(2,2))
plot(rent_causal_mod)
par(mfrow = c(1,1))
```

There are likely some issues that show up in the diagnostics.

+ The first is unequal variance of errors. If you think about it, it's maybe not too surprising for two reasons.

  - In your EDA, if you did a histogram of `Price`, then you may have noticed that rent tends to be rounded to £25-£50 increments when the rent is less than £1000, but when the property is more than £1000, this rounding increases to £50-£100. This creates "error" in the price measurement at higher prices, so our model uncertainty performs "worse".
  
  - Keep in mind that unequal variance of errors impacts our estimated standard error, which in turn impacts our 95% confidence intervals. It does not affect our actual prediction.
  
  - The other reason is that estate agents have a easier time guessing the rent for "cheap" properties. For cheap flats, it’s easier to pin down the right rent because £250 is a big chunk of the total (e.g. going from £500 to £750 is a big jump). For a luxury mansion, £250 hardly moves the needle (e.g. going from £3,500 to £3,750), so estimates naturally spread out more as price increases.
  
+ The second issue you might see is that you may have one or multiple overly influential observations. This has fluctuated over time as I've added more and more data (sometimes these disappear, other times more appear) but generally there are one or two expensive properties that cause issues and need to be addressed.

  - Influential observations do impact the parameter estimates, so you would first want to confirm if the entered `Price` is correct (which you can check given you have `PropertyURL` - but note that ASPC sometimes removes properties)
  
  - If you confirm the `Price` has been recorded properly, then you will need to decide how to resolve. Generally, you have two options, either ignore the problem (but acknowledge it when talking about your results) or remove the observation from your dataset. There are some other options, but we won't go into them here.
  
Fix any issues with influential properties, and rerun your analysis.

```{r, include=answer_sheet, class.source="bg-success", eval = T}
# If your row numbers are different to mine, make sure you re-loaded the data in before starting with the causal model and note that by the time of this workshop (and with more data) observation 73 may not be a problem anymore, or it may have changed to a different observation.
abdn_trimmed <- abdn[-c(73),]
rent_causal_mod <- lm(Price ~ FloorArea, data = abdn_trimmed)
# Diagnostics now look good except for unequal variance of errors.
```

## 6. Summarise the results.

The technical steps are exactly the same as for the predictive model. Get the model summary:

```{r}
summary(rent_causal_mod)
```

```{r, echo = F}
beta_dag <- unname(coef(rent_causal_mod)["FloorArea"])
```

And plot:

```{r}
fake <- data.frame(
  FloorArea = seq(from = min(abdn$FloorArea), to = max(abdn$FloorArea), length.out = 10)
)
preds <- predict(rent_causal_mod, newdata = fake, se.fit = TRUE)
fake$fit <- preds$fit
fake$low <- preds$fit - preds$se.fit * 1.96
fake$upp <- preds$fit + preds$se.fit * 1.96
ggplot(fake) +
  geom_ribbon(aes(x = FloorArea, ymin = low, ymax = upp), fill = "lightgrey", alpha = 0.7) +
  geom_line(aes(x = FloorArea, y = fit)) +
  labs(x = "Square meters of property",
       y = "Predicted rent") +
  theme_bw()
```

So what we learn is that for every additional meter square "added" the rent increases by roughly `r gbp(beta_dag)` per month. So long as we assume that our DAG is accurate (a potentially big assumption), this `r gbp(beta_dag)` increase is the *causal* effect of square meters on rent.

Compare that with our predictive model that found "adding" an extra square meter to a property would increase the rent by `r gbp(beta_mod)` per month. At the time, we noted that `r gbp(beta_mod)` seemed on the "cheap" side and the reason for the predictive modelling estimating it as being "cheap" was because, according to out DAG, m$^2$ was confounded by the number of rooms in a flat. To be clear, that confounding didn't matter for the first model because our aim was to make predictions, not to understand the causal effect.

# What is the causal effect of EPC on Price?

Repeat the above process, but where our research question is focused on understanding the causal effect of EPC on Price.

Because we're doing a new analysis, be sure to re-load the data before proceeding.

```{r, include=answer_sheet, class.source="bg-success", eval = F}
# This is the same DAG as above. Nothing actually changes in the DAG even though our research question has.
rent_dag <- check_dag(
  Price ~ Rooms + FloorArea + Tax + EPC,
  Rooms ~ FloorArea,
  EPC ~ FloorArea,
  Tax ~ FloorArea,
  exposure = "EPC",
  outcome = "Price"
)

rent_dag
# To estimate the causal effect of EPC, we need to include FloorArea in the model:
EPC_mod <- lm(Price ~ EPC + FloorArea, data = abdn)
# Check diagnostics:
plot(EPC_mod) # No issues with this model
summary(EPC_mod)
plot(ggpredict(EPC_mod, terms = "EPC"))
# The causal effects seem to be all over the place. This either suggests that the DAG may be more complex than I've assumed, or we need more data to estimate this relationship properly.

# Note that we do not interpret FloorArea in this model. FloorArea is included simply to deal with confounded relationships, so under this version of the model is it now confounded. If we did interpret FloorArea from this model we would be falling into the "Table 2 Fallacy".

# As an example of how drawing a DAG can be challenging, below is a more fleshed out version of the DAG that has had more thought given to it:
better_dag <- check_dag(
  Price ~ Rooms + FloorArea + Tax + EPC + PropertyType,
  Rooms ~ FloorArea + PropertyAge,
  EPC ~ FloorArea + PropertyAge,
  Tax ~ HistoricPrice,
  HistoricPrice ~ PropertyAge + PropertyType + Rooms + FloorArea,
  FloorArea ~ PropertyAge + PropertyType,
  exposure = "EPC",
  outcome = "Price"
)
# (But keep in mind, this doesn't include things like how nicely it's decorated, it's general maintainence, how greedy the landlord is, etc, etc)
plot(better_dag)
better_dag
# Which tells us that we'd one of the following models:
# Rent ~ EPC + PropertyType + Rooms + FloorArea + Tax
# Rent ~ EPC + HistoricPrice + PropertyType + Rooms + FloorArea
# Rent ~ EPC + PropertyAge + FloorArea

# The easiest of which would be the first, given PropertyType is the easiest variable to collect (property age is difficult, and historic Price may well beimpossible to get). The means that unless we get PropertyType, we can't currently understand the causal effect of EPC with the data as is. This is a nice example of where doing the DAG *before* data collection helps to prevent headaches later on by identifying which variables to collect!
```

# Assumption table

| **Assumption**                   | **Description**                                                                                                      | **Example of Violation**                                                                                                                                                        |
|---------------|-----------------------|-----------------------------------|
| 1\. **Validity**                 | The data is able to answer the research question effectively.                                                        | Using IQ to measure overall intelligence (what does IQ even measure?).                                                                                                          |
| 2\. **Representativeness**       | The data reflects the population or group being studied.                                                             | Using survey data from a rural area to draw conclusions about an urban population.                                                                                              |
| 3\. **Additivity**               | Effects of predictors are additive and there are no interactions between them.                                       | Assuming that study hours and sleep hours independently affect exam scores without considering their combined effect (imagine a Venn diagram with study and sleep overlapping). |
| 4\. **Linearity**                | The relationship between predictors and the outcome is linear.                                                       | Assuming plant growth will continue to increase indefinitely without ever levelling out.                                                                                        |
| 5\. **Independence of Errors**   | Observations are independent of each other. (Related to residual errors.)                                            | If modelling height, the data come from the same group of children over time.                                                                                                   |
| 6\. **Equal Variance of Errors** | Residuals have constant variance across all levels of predictors ("homoscedasticity"). (Related to residual errors.) | If modelling income, variability in salaries is not the same across all levels of a company's hierarchy.                                                                        |
| 7\. **Normality of Errors**      | Residuals should be normally distributed. (Related to residual errors.)                                              | If modelling income, the presence of any extremely wealthy individuals will be hard for the model to understand and explain.                                                    |
