---
title: "Linear models with single cagegorical predictor - Study Notes"
author: "Michelle Pinard & Deon Roos"
date: 'Last updated: `r Sys.Date()`'
output:
  word_document:
    toc: yes
    reference_docx: "../sbstemplate.docx"
---

# Purpose

When the explanatory variable defines groups, we model differences in group means rather than a straight-line trend. The model answers whether average outcomes differ between groups and by how much.

# How to read the model

Linear models with a categorical predictor estimate a reference-group mean (the intercept) and one contrast per additional group. Each contrast is the mean difference relative to the reference. Report both the estimate and its uncertainty.

# How a categorical predictor is encoded

For the model `lm(y ~ group, data = df)`, where `df$group` has the levels `A`, `B` and `C`, `A` will be taken as the reference group, and become the intercept, since it appears first alphanumerically. For group `B` and `C`, two "temporary columns" are added to the dataset, whereby a value of 1 is given if it belongs to that specific group. We do not need a "temporary column" for `A` because it is the intercept (i.e. if we know an observation is not `B` or `C`, then it must be `A`).

Here's what it looks like for a simple example:

```{r, echo = F, eval = T}
library(knitr)
kable(data.frame(
  response = round(rnorm(n = 6, 10, 2), digits = 1),
  group = c("A", "A", "B", "C", "B", "C"),
  B = c(0,0,1,0,1,0),
  C = c(0, 0, 0, 1, 0, 1)
))
```

This allows us to use the following model:

$y_i \sim Normal(\mu_i, \sigma)$

$\mu_i = \beta_0 + \beta_1 \times B_i + \beta_2 \times C_i$

* $\beta_0$ is the intercept, or average for group `A`

* $\beta_1$ is the difference between group `A` and group `B` and is multiplied with $B$ which is now either 1 or 0 ("on" or "off")

* $\beta_2$ is the difference between group `A` and group `C`, and is multiplied with $C$ which is either 1 or 0 ("on" or "off")

Doing so allows us to include a category into the model.

# Uncertainty to report

95% confidence intervals (CI) are a means to convey uncertainty around estimates. Before fitting, the CI _procedure_ has a 95% chance to "hit the target" (i.e. hit the true value) but after fitting, the interval either covers the true value or not. 

When interpreting, consider the point estimate and its CI together. Avoid treating “significance” as the only decision rule.

CI can be calculated using estimated standard errors for each parameter multiplied by 1.96 (a critical value) and either subtracted or added to the parameter estimate.

# Workflow carried across

Keep the usual sequence: formulate a question, do quick EDA, note hidden assumptions, fit the model, diagnose (not done in these Study Notes but the next), summarise, then interpret with the stated caveats. 

Hidden assumptions include validity and representativeness, which are not checked by residual plots and must be discussed explicitly.

# Using categories wisely

Categorising a continuous variable multiplies the number of parameters needed to fit the model and can massively reduce efficiency. Sometimes categories are practical or appropriate but most of the time a continuous measurement is more informative.

# Useful code examples

```{r, eval = F}
# Fitting the model
m <- lm(y ~ group, data = d)
# Get summary of model
summary(m)

# Generate predictions to show relationship
fake <- data.frame(group = unique(d$group))
pr   <- predict(m, newdata = fake, se.fit = TRUE)

# Calculate 95% CI
fake$fit <- pr$fit
fake$low <- pr$fit - 1.96 * pr$se.fit
fake$upp <- pr$fit + 1.96 * pr$se.fit

# Plot relationship
ggplot(fake, aes(group, fit)) +
  geom_errorbar(aes(ymin = low, ymax = upp), width = 0.15) +
  geom_point()

# Alterantively, use
library(ggeffects)
preds <- ggpredict(m)
plot(preds)

# To change a character to a factor
d$group <- factor(d$group)

# To set the reference level of a category:
d$group <- factor(d$group, levels = c("B", "A", "C"))
# Or
d$group <- relevel(df$group, ref = "A")
```

