---
title: "Workshop 1: Getting Back into R"
subtitle: "BI3010 Statistics for Biologists"
author: "Roos & Pinard"
date: last-modified
date-format: "D MMMM YYYY"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    theme: [cosmo, ../custom.scss]
    highlight-style: monokai
    embed-resources: true
    code-copy: true
    code-tools: false
    html-math-method: plain
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
ali   <- read.table("data/alligator_bite.txt", header = TRUE)
fast  <- read.table("data/animal_speed.txt",   header = TRUE)
chupa <- read.table("data/chupa.txt",          header = TRUE)

make_download_btn <- function(path, label = NULL) {
  fname <- basename(path)
  if (is.null(label)) label <- paste("Download", fname)
  uri <- xfun::base64_uri(path)
  sprintf('<a href="%s" download="%s" class="ws-download-btn">%s</a>', uri, fname, label)
}
```

```{=html}
<script>
function revealAnswer(btn) {
  var block = btn.closest('.question-block, .puzzle-block');
  var answerClass = block.classList.contains('puzzle-block') ? '.puzzle-answer' : '.instructor-answer';

  var textarea = block.querySelector('.student-answer');
  if (textarea) {
    if (textarea.value.trim() === '') {
      textarea.classList.add('missing');
      textarea.placeholder = 'Please write your answer before revealing...';
      textarea.focus();
      return;
    }
    textarea.classList.remove('missing');
    textarea.disabled = true;
    textarea.classList.add('ws-answer-locked');
    var badge = document.createElement('span');
    badge.className = 'ws-submitted-badge';
    badge.textContent = 'Submitted';
    textarea.insertAdjacentElement('afterend', badge);
  }

  var mcOptions = block.querySelector('.mc-options');
  if (mcOptions) {
    var radios = mcOptions.querySelectorAll('input[type=radio]');
    var anyChecked = Array.from(radios).some(function(r) { return r.checked; });
    if (!anyChecked) {
      mcOptions.classList.add('unanswered');
      return;
    }
    mcOptions.classList.remove('unanswered');
    radios.forEach(function(r) {
      r.disabled = true;
      var label = r.closest('label');
      if (r.dataset.correct === 'true') {
        label.classList.add('mc-correct');
      } else if (r.checked) {
        label.classList.add('mc-incorrect');
      }
    });
  }

  var answer = block.querySelector(answerClass);
  if (answer) answer.style.display = 'block';
  btn.textContent = 'Answer shown';
  btn.disabled = true;
}

function toggleGlossary() {
  document.getElementById('glossary-panel').classList.toggle('open');
}
</script>
```

```{=html}
<style>
#glossary-btn {
  position: fixed;
  bottom: 2rem;
  right: 2rem;
  background: #00b494;
  color: #fff;
  border: none;
  padding: 0.55rem 1.1rem;
  border-radius: 4px;
  cursor: pointer;
  font-size: 0.875rem;
  font-weight: 600;
  z-index: 1000;
  box-shadow: 0 2px 8px rgba(0,0,0,0.25);
  transition: background 0.15s;
  font-family: inherit;
}
#glossary-btn:hover { background: #009478; }

#glossary-panel {
  position: fixed;
  top: 0;
  right: -400px;
  width: 380px;
  height: 100vh;
  background: #fff;
  border-left: 3px solid #00b494;
  box-shadow: -4px 0 20px rgba(0,0,0,0.15);
  z-index: 1001;
  transition: right 0.3s ease;
  display: flex;
  flex-direction: column;
}
#glossary-panel.open { right: 0; }

#glossary-header {
  background: #0d1e2c;
  color: #fff;
  padding: 1rem 1.25rem;
  display: flex;
  justify-content: space-between;
  align-items: center;
  flex-shrink: 0;
}
#glossary-header span {
  font-weight: 600;
  font-size: 0.95rem;
  letter-spacing: 0.03em;
}
#glossary-close {
  background: none;
  border: none;
  color: #fff;
  font-size: 1.3rem;
  cursor: pointer;
  padding: 0;
  line-height: 1;
  opacity: 0.7;
}
#glossary-close:hover { opacity: 1; }

#glossary-body {
  overflow-y: auto;
  flex: 1;
}

.glossary-entry {
  padding: 0.85rem 1.25rem;
  border-bottom: 1px solid #e8f5f0;
}
.glossary-entry:last-child { border-bottom: none; }

.glossary-terms {
  display: flex;
  align-items: baseline;
  gap: 0.5rem;
  flex-wrap: wrap;
  margin-bottom: 0.3rem;
}
.glossary-en-term {
  font-weight: 600;
  color: #0d1e2c;
  font-size: 0.92rem;
}
.glossary-zh-term {
  color: #00b494;
  font-weight: 600;
  font-size: 0.92rem;
}
.glossary-en-def {
  font-size: 0.845rem;
  color: #444;
  line-height: 1.5;
  margin-bottom: 0.25rem;
}
.glossary-zh-def {
  font-size: 0.845rem;
  color: #1c3a50;
  line-height: 1.65;
}

.ws-answer-locked { background: #f5f5f5 !important; color: #555 !important; cursor: not-allowed; }
.ws-submitted-badge { display: inline-block; margin-top: 0.3rem; background: #00b494; color: #fff; font-size: 0.75rem; font-weight: 600; padding: 0.2rem 0.55rem; border-radius: 3px; }
.glossary-section-header {
  padding: 0.4rem 1.25rem 0.3rem;
  font-size: 0.72rem;
  font-weight: 700;
  text-transform: uppercase;
  letter-spacing: 0.07em;
  color: #00635a;
  background: #f0faf7;
  border-top: 2px solid #e8f5f0;
  margin-top: 0.25rem;
}
.glossary-section-header:first-child { border-top: none; margin-top: 0; }
.ws-download-btn { display: inline-block; margin: 0.4rem 0; padding: 0.35rem 0.9rem; background: #0d6efd; color: #fff !important; border-radius: 4px; text-decoration: none !important; font-size: 0.875rem; font-weight: 500; }
.ws-download-btn:hover { background: #0b5ed7; }
</style>
```

Welcome to the first BI3010 workshop. Over the next ten workshop sessions, you'll go from working with raw data to fitting complex statistical models, building up a workflow you can apply to almost any dataset you encounter, including every single one of your honour's project next year. Today's workshop focuses on getting comfortable with R and learning to spot problems in data before analysis begins, a step that sounds unglamorous but is where scientists spend 80% of their time.

# Learning Objectives

::: {.callout-tip}
## By the end of this workshop you will be able to:

1. (Re)familiarise yourself with the basics of `R`;
2. Use `R` to perform simple descriptive summaries of data and data processing;
3. Use the results of the EDA (Exploratory Data Analysis) to identify any issues in the data;
4. Begin thinking about the assumptions we make with a specific dataset even before we begin using statistics.

All of these objectives link to useful skills for **initial, quick explorations** and screening of datasets. Because it's so important to ensure that the data we subsequently use in analysis is suitable for answering our question, the importance of such EDA cannot be overstated.

These skills will be used throughout the course.
:::

# Workshop structure

::: {.callout-note}
## The Analytical Workflow

The workshops on this course will have you learn an analytical workflow, focussing mainly on analysing data using linear models (and eventually generalised linear models). That is, you're going to learn how to go from getting a dataset to extracting evidence to answer your research question. Importantly though, this workflow is not unique to linear models (or GLMs). You would (or at least *should*) use this same general workflow, regardless the dataset, the scenario, or the analytical method you may find yourself using.

The general workflow we'll be using can be broken down into:

1. Formulate a research question.
2. Perform exploratory data analysis [Focus of today].
3. Identify any hidden assumptions.
4. Fit an appropriate model.
5. Diagnose the model and check assumptions.
6. Summarise the results.
7. Interpret and provide inferences.

For the BI3010 workshops we are generally going to skip Step 1 in the workflow for convenience sake, but in your own work Step 1 is key as every subsequent step in the workflow is dependent on your question. In your own work (e.g. your Honour's project) it's important to take the time to think carefully about what the question you want to answer is, how best to design an experiment or data collection protocol to get data suitable for answering it, which statistical method would be suitable for analysing the data, and what caveats and limitations come with all of these. Basically, don't take the omission of Step 1 in this course as evidence that it's not important. It is.
:::

# Getting Started with R

::: {.callout-note collapse="true"}
## New to RStudio? Expand here

Before we begin the workshop, let's make sure that you're set up to work with `R` and RStudio. If you're working on your own machine, make sure you have `R` and RStudio installed. If you don't, you can download `R` from [CRAN](https://cran.r-project.org/) and RStudio from [RStudio's website](https://www.rstudio.com/products/rstudio/download/). On university computers, these should already be installed.

Once you have `R` and RStudio set up, open RStudio. You should see a window with four main panes:

- **Source:** This is where you write your scripts.
- **Console:** This is where you can type and execute R code directly.
- **Environment/History:** This pane shows you the objects that are currently stored in R's memory.
- **Files/Plots/Packages/Help/Viewer:** This is a multi-tabbed pane that allows you to navigate files, view plots, manage packages, and access documentation.

**Running Code**

There are a few ways to run `R` code:

- **Directly in the console:** You can type `R` commands directly into the Console pane and press Enter to run them. This is useful for very quick tests and calculations but avoid using this for project work (none of it's saved).
- **From a script:** Scripts are simple text files that contain `R` code. You can write multiple lines of code in a script and run them together or one by one. This is particularly useful for more complex analyses where you want to keep a record of your code.

To run code from a script, you can:

- Click the "Run" button at the top of the Source pane to run the selected line or code chunk.
- Press `Ctrl + Enter` (or `Cmd + Enter` on a Mac) to run the current line or selected code.
:::

# Instruction Orientation

## Commenting guide

It's good practice to make comments to yourself (and others) while writing code. Comments generally act as notes that help explain your general intent for a section of code, which in three weeks' time can be an absolute lifesaver. They're also super useful for anyone else reading your code (and your commenting may be evaluated for some jobs).

There are many different opinions as to how to write good comments (see [here](https://statmodeling.stat.columbia.edu/2024/02/07/when-all-else-fails-add-a-code-comment/) for a particularly stern view), but a general principle I subscribe to is: aim to very concisely describe your general purpose for the next few lines of code. What are *you* trying to do, and not, what does *each line of code* do.

For example, I would write:

```{r, eval=FALSE}
# Get R set up
library(ggplot2)  # For plots
setwd("my/file/path")
my_data <- read.csv("data/sales_data.csv")

# Plot the data
ggplot(my_data, aes(x = country, y = total_sales)) +
  geom_bar()
```

And not:

```{r, eval=FALSE}
# Load the ggplot2 package into the R environment so we can create plots later
# What happened to ggplot1?
library(ggplot2)

# Use the read.csv function to read the sales_data.csv file from the data directory
# and store the resulting data frame in a data frame called 'my_data'
# Remember that it's called my_data and not data. Calling it data made things break :(
my_data <- read.csv("data/sales_data.csv")
# my_data <- read.table("data/sales_data.csv") # DONT USE THIS IT DOESNT WORK
# mi_data <- read_csv("data/sales_data.csv") # ALSO DIDN'T WORK - DUNNO WHY

# Use ggplot to create a bar plot of the total sales by country for 2024.
# The x-axis represents the categories, and the y-axis represents the total sales.
ggplot(my_data, aes(x = country, y = total_sales)) + # using ggplot, take my_data, then plot country on the x and sales on the y
  geom_bar() # then have the data visualised as bar, using the structure described in the previous line of code
# When I see comments like this IRL, I die a little bit inside. Keep it clean and concise.
```

The latter is so overly descriptive that it becomes as much of a chore to read the comments as the actual code itself. In fact, I think the code does a better job of quickly and efficiently explaining its purpose than these comments do! Another way to think of how to best write comments is: think of comments as post-it notes. You want them to be short and to act as a signal for sections that you want to dive deeper into (e.g. for debugging).

## On using LLMs

::: {.callout-caution}
## On using LLMs

LLMs (Large Language Models, or AI) are now commonly used within programming and data analysis. They can be good for boring or fiddly tasks, one-off problems, and tidying code (see [here](https://nicholas.carlini.com/writing/2024/how-i-use-ai.html) for a pragmatic take on this). The catch is that you need to know enough about programming to tell when the code they produce is correct and when it's plausible-sounding nonsense. That line blurs fast as the code gets more complex, especially when the code is more complex than you are capable of producing yourself. For analysis specifically, without a sound theoretical understanding of statistics and data, they tend to get it wrong **60%** of the time if the user can't specify their data and question properly (see [here](https://doi.org/10.1111/2041-210x.70267)). To make sure that doesn't happen to you, you need to understand statistical theory and assumptions.

Within BI3010, you are allowed to use an LLM (or genAI) during workshops but *not during tests*. I would advise using them with caution though. Keep in mind you are here to learn. Offloading that task to a LLM means you aren't learning. I'd suggest you try to solve any problems yourself first. Make a mistake and then try to learn from the mistake. If you can't solve it, you are welcome to have an LLM try help to solve the problem. If it does solve the problem, try to understand how it solved it.

Using them effectively and responsibly is a skill in itself. Make your prompt specific: describe what you want, copy and paste the relevant code that gives an error, include any error messages, and say you're working in `R` (otherwise you might get Python thrown back at you). Mention any packages you want to use, or specify base `R`.
:::

# Workshop Instructions

In this workshop, we're going to load two datasets, starting with alligator bite strength, and carry out Exploratory Data Analysis (EDA) along with some simple data quality assurance. Broadly, our aim in such cases is to identify *weird* things in the data that might indicate mistakes, such as data entry errors. In real-world analysis, this step is incredibly important because mistakes are easily made, especially during data entry. Imagine entering 5,000 nearly identical observations into a spreadsheet. People get bored, start thinking about other things, get mixed up about which observation they should be adding, among many other distractions. It's hard to stay super focused during mundane work for long periods. It would be unrealistic to expect or assume that mistakes would never happen, so we need to check for them. The problem is that if we don't spot these errors, they can significantly influence the results when it comes time to analyse the data.

The good news is that we don't need any complex statistics to check for these issues, and the `R` coding is *relatively* straightforward.

## Setting Up Your Project

We'll use an R Project to organise our work. An R Project is just a folder somewhere on your PC with a small `.Rproj` file inside it. When you open that file, RStudio automatically sets the working directory to the project folder (basically saying "go here to find the data and scripts"), with no manual path-setting required, and it will work the same on any computer or operating system.

**Step 1: Create the folder structure**

Create a folder called `BI3010` somewhere sensible:

- University computers: the H drive works well, e.g. `H:\BI3010`
- Personal computers: anywhere in your Documents folder is fine

Inside `BI3010`, create two subfolders:

- `data`: all data files for the course go here
- `scripts`: your R scripts, one per workshop

Your folder should look like this:

```
BI3010/
├── data/
└── scripts/
```

**Step 2: Create the R Project**

Open RStudio and go to `File > New Project > Existing Directory`. Navigate to your `BI3010` folder and click `Create Project`. RStudio will create a `BI3010.Rproj` file inside the folder and restart with the project open.

From now on, you can either open RStudio by double-clicking `BI3010.Rproj` or by opening RStudio directly and using the `Project` drop down button at the top right to select `BI3010.Rproj`. This ensures the working directory is always set correctly.

**Step 3: Create a script for this workshop**

Go to `File > New File > R Script` (or `Ctrl + Shift + N`). Save it immediately into your `scripts` folder as `workshop1.R`. To run code, press the Run button or use `Ctrl + Enter`.

## Download the Data

The data file for this section is embedded in this page. Click the button below to download it, then save it to your `BI3010/data/` folder before continuing.

```{r, echo=FALSE, results='asis'}
cat(make_download_btn("data/alligator_bite.txt"))
```

## Load File into RStudio

Now we can load the data. The `.txt` extension tells us it's a plain text file, so we use `read.table()`. We need two arguments: the filename (including the `data/` subfolder path), and `header = TRUE` to tell `R` that the first row contains column names rather than data.

Because the project sets the working directory to `BI3010/`, we can refer to files using "relative paths" (e.g. `data/alligator_bite.txt`) rather than "absolute paths" (e.g. `H:/BI3010/data/alligator_bite.txt`). Doing this saves yourself a bit of typing but also means you can share the entire folder with someone else and they don't need to change the path to where ever they saved the project.

Tip: `?read.table` (or `?` followed by any function name) opens the help page for that function. For example, you can use this to figure out what `header = TRUE` is doing in the `read.table()` code below.

```{r, eval=FALSE}
ali <- read.table("data/alligator_bite.txt", header = TRUE)
```

The above fits on one line, but you can split it for readability:

```{r, eval=FALSE}
ali <- read.table("data/alligator_bite.txt",
                  header = TRUE)
```

## Checking It Worked

If you hit an error when trying to run code, don't panic. Nothing in `R` is permanent (unless you very explicitly make it so). If you hit an error, read the error message, trouble shoot solutions, run the code again, and move on (assuming it was fixed). The only thing that's permanent is your script and original data, so save your script often and write comments as you go. If you think you might have some redundant code that you're not sure whether removing it will break your entire script, then comment it out with `#` first rather than deleting it (remember that comments are not read as code).

If you get an error loading data, common causes are:

- The file isn't in the `data` folder inside your `BI3010` project, or you didn't open RStudio via `BI3010.Rproj`.
- You may have misspelled one of the words (including using a capital letter instead of lower case).
- Your file may be missing a column heading.
- The file isn't in the correct format (e.g., the file saved as a `.csv` instead of a `.txt`).
- The quotation marks are in the wrong location.
- You may have missed the final bracket `)` in the `read.table()` function.

To confirm that the file has been read properly, you should see it in the Environment panel (normally in the upper right of RStudio).

To do an initial quick inspect the data, you have a few options. The following shows the first six rows (use `n` to change how many, e.g. `head(ali, n = 10)`):

```{r, results="hide"}
head(ali)
```

Or open the full dataset in a new tab; `View()` gives a spreadsheet-like view with sorting and filtering:

```{r, eval=FALSE}
View(ali)
```

Or print everything to the console:

```{r, eval=FALSE}
ali
```

Not recommended: it floods the console and is hard to read.

## Question Set One

<div class="question-block">
<p><strong>Question 1.</strong> How many variables do we have in this dataset?</p>
<div class="mc-options">
  <label><input type="radio" name="q1s1_q1"> 1</label>
  <label><input type="radio" name="q1s1_q1" data-correct="true"> 2</label>
  <label><input type="radio" name="q1s1_q1"> 4</label>
  <label><input type="radio" name="q1s1_q1"> 6</label>
</div>
<button class="reveal-btn" onclick="revealAnswer(this)">Show Answer</button>
<div class="instructor-answer">
<p>There are <strong>2 variables</strong>. There are several ways to check this:</p>
<pre><code class="language-r">summary(ali)
ncol(ali)   # Number of columns (i.e., variables) in the ali dataset
str(ali)    # Structure of the ali dataset
head(ali)   # Displays the first few rows
View(ali)   # Opens the dataset in a spreadsheet-like view</code></pre>
</div>
</div>

<div class="question-block">
<p><strong>Question 2.</strong> What are the variable names?</p>
<div class="mc-options">
  <label><input type="radio" name="q1s1_q2"> <code>size</code> and <code>force</code></label>
  <label><input type="radio" name="q1s1_q2" data-correct="true"> <code>length</code> and <code>bite</code></label>
  <label><input type="radio" name="q1s1_q2"> <code>length</code> and <code>pressure</code></label>
  <label><input type="radio" name="q1s1_q2"> <code>weight</code> and <code>bite</code></label>
</div>
<button class="reveal-btn" onclick="revealAnswer(this)">Show Answer</button>
<div class="instructor-answer">
<p>The variable names are <code>bite</code> and <code>length</code>. The functions above will show you this (except <code>ncol()</code>, which only counts columns).</p>
</div>
</div>

<div class="question-block">
<p><strong>Question 3.</strong> Are the variable names capitalised?</p>
<div class="mc-options">
  <label><input type="radio" name="q1s1_q3" data-correct="true"> No, both are lower case</label>
  <label><input type="radio" name="q1s1_q3"> Yes, both start with a capital letter</label>
  <label><input type="radio" name="q1s1_q3"> Only <code>Length</code> is capitalised</label>
  <label><input type="radio" name="q1s1_q3"> Only <code>Bite</code> is capitalised</label>
</div>
<button class="reveal-btn" onclick="revealAnswer(this)">Show Answer</button>
<div class="instructor-answer">
<p>No, <code>bite</code> and <code>length</code> are both lower case. A general tip: keep variable names short and all lower case. You will type them a lot, so save yourself the effort. Avoid names like <code>Alligator_Bite_Strength_PSI</code>.</p>
</div>
</div>

These questions may seem trivial, but honestly, they're the first questions we ask ourselves when we load a dataset into `R` for the first time.

# Getting to Know the Data

Here's what we know about the data. The data contains `r nrow(ali)` observations (sometimes referred to as samples) of alligator bite strength, measured in pounds per square inch. All `r nrow(ali)` alligators were measured at Parker Island Gator Farm, Florida, USA. For each bite strength, the length of the alligator, from nose to tail, was measured and is provided in centimeters (to one decimal place).

Before we move ahead with the data, let's take a second to become familiar with alligator bite strength. Using your search engine of preference, do a quick search to find out the bite strength of alligators. Do the same for the length of an average alligator.

When we do exploratory data analysis, we're specifically looking for things that look *weird*. In order to know if something is *weird*, we need to know what "normal" is; hence why it's a good idea to have some sort of notion of what a "normal" alligator's bite strength and length may be.

Write down the average bite strength of an alligator, measured in pounds per square inch, and the average length of an alligator in centimeters (ideally as a comment in your script).

With that done, let's move on to understanding our data. For starters, let's check if the data is sensibly structured. In `R`, this means that numbers are numbers, words are characters (or sometimes factors), and so on. Basically, we want to make sure there aren't any bite measurements like `2174.94BIGBITE`, which would mess us up by causing `R` to think that all `bite` values are characters and not numbers.

To do this quickly, we can use the `str()` function (short for "structure"):

```{r, results="hide"}
str(ali)
```

We see that `ali` is a `data.frame` (great, as it should be!), with `r nrow(ali)` observations (or samples) and two variables. The `$` operator is used to extract specific variables from a dataframe, so `ali$bite` will give you just the `bite` variable from the `ali` dataset. From this, we can see that both the `length` and `bite` variables are being treated as numbers (indicated by `num`), and we get a handful of example values for both `length` and `bite`.

So far, we're not seeing anything unusual, which is great, it means we haven't identified any problems that we need to fix.

If we wanted to, we could extract the `length` variable from the dataset by doing:

```{r, results="hide"}
ali$length
```

How would you extract the `bite` variable?

Let's use these extracted variables to calculate some summary statistics, starting with the mean (i.e. average). To do so, we use the `mean()` function. To find the average length of our alligators, we do:

```{r, results="hide"}
mean(ali$length)
```

We can see that the average alligator in Parker Island Gator Farm is `r round(mean(ali$length), digits = 1)` cm long.

Compare this value with the mean length of alligators you found from a quick internet search. It might be a bit longer than expected, but this could be due to these alligators coming from a farm rather than the wild. Perhaps they're given lots of food to help them grow? While our numbers may not match exactly with online data, significant discrepancies could be worth investigating.

Similarly, calculate the average bite strength of our alligators and see if there's anything unusual there.

```{r, results="hide"}
mean(ali$bite)
```

Assuming you found that alligator bite strength is roughly 2,000 pounds per square inch, our alligators appear to be biting with an extra 1,500 pounds per square inch *on average*. That seems like *a lot* more.

This is something that should make us sit up and pay attention. We don't know for sure if there's a problem, but we have reason to believe there *might* be one. It's now up to us to prove that there *isn't* a problem.

There are many descriptive techniques to check for anomalies in our data, but a convenient "short cut" is to use the `summary()` function. This function provides a generic summary of our entire dataset when we supply `ali` as the sole argument. Try this now and see if you can identify any issues with the data:

```{r, results="hide"}
summary(ali)
```

Keep in mind, with EDA we're looking for anything *weird* in the data.

## Question Set Two

<div class="question-block">
<p><strong>Question 1.</strong> How small is the smallest alligator in the farm for which we have data, and what's the biggest?</p>
<textarea class="student-answer" rows="2" placeholder="Write your answer here..."></textarea>
<button class="reveal-btn" onclick="revealAnswer(this)">Show Answer</button>
<div class="instructor-answer">
<p>The smallest alligator is <strong>153.5 cm</strong> and the largest is <strong>2293.0 cm</strong>. Multiple approaches work:</p>
<pre><code class="language-r">range(ali&#36;length)    # minimum and maximum
min(ali&#36;length)      # minimum
max(ali&#36;length)      # maximum
summary(ali)         # summary for all variables
summary(ali&#36;length)  # summary for length only</code></pre>
</div>
</div>

<div class="question-block">
<p><strong>Question 2.</strong> What are the minimum and maximum values for bite strength?</p>
<textarea class="student-answer" rows="2" placeholder="Write your answer here..."></textarea>
<button class="reveal-btn" onclick="revealAnswer(this)">Show Answer</button>
<div class="instructor-answer">
<p>The minimum bite strength is <strong>832.9 PSI</strong> and the maximum is <strong>123,632.0 PSI</strong>.</p>
<pre><code class="language-r">range(ali&#36;bite)</code></pre>
</div>
</div>

<div class="question-block">
<p><strong>Question 3.</strong> Do any of these minimums or maximums seem strange?</p>
<div class="mc-options">
  <label><input type="radio" name="q2_q3" data-correct="true"> Yes, both the maximum length and maximum bite strength look suspicious</label>
  <label><input type="radio" name="q2_q3"> Yes, only the maximum length looks suspicious</label>
  <label><input type="radio" name="q2_q3"> Yes, only the maximum bite strength looks suspicious</label>
  <label><input type="radio" name="q2_q3"> No, all values seem reasonable</label>
</div>
<button class="reveal-btn" onclick="revealAnswer(this)">Show Answer</button>
<div class="instructor-answer">
<p>Yes. A 2293 cm alligator is longer than a bus. And 123,632 PSI is about 60 times the average bite strength of 2,000 PSI. At least one of these values is almost certainly a data entry error.</p>
</div>
</div>

# Data Detective

Hopefully, you've identified at least one questionable value in our dataset. What should we do now? Throw out the entire dataset? (No. The answer is "No, you do not throw out the dataset", at least not yet.) Let's try to figure out what's going on, starting with finding the problem observations.

How do we do that?

Let's start by considering what we think a weird observation might look like. We can use the `summary(ali)` output to help us here. For instance, we know we have an alligator that was apparently `r max(ali$length)` cm long (`r round(max(ali$length)/100, digits = 0)` m, that's longer than a bus!). How do we find that value in our dataset to see if there was anything else unusual about it?

This is where square brackets `[]` come into play. In `R`, square brackets are used for "indexing" to extract specific rows or columns from a dataframe. Think of square brackets as a way to point to or select parts of your data.

To use them effectively, first consider how you would find the information manually (e.g. in Excel):

1. We'd start by looking at the `ali` dataset.
2. We'd then skim through the `ali` dataset, looking for any rows where alligators are `r max(ali$length)` cm long.
3. Once we've found these rows, examine all the information related to those alligators.

How does that translate into `R`? Almost exactly; we just need to write those three steps using code.

First, we'd start by looking at the `ali` dataset.

```{r, eval=FALSE}
ali
```

Next, we'd skim through the `ali` dataset, looking for any rows where alligators are `r max(ali$length)` cm long. This is where we can use square brackets `[]` in conjunction with conditional statements. Conditional statements allow us to specify criteria like "when this condition is met, do *something*".

In our case, the condition is finding alligators that are `r max(ali$length)` cm long. In `R`, we write this condition using double equal signs (`==`) to denote "equal to" (a single `=` assigns a value; `==` tests whether two things match). This is basically the same as your eyes going down the Excel column looking for `r max(ali$length)`, noting any instances where it was true or not. For example:

```{r, results="hide"}
ali$length == 2293.0
```

The output is a list of `r nrow(ali)` values, each being either `TRUE` or `FALSE`, corresponding to each row in the `ali` dataset. `TRUE` values indicate rows where the alligator is `r max(ali$length)` cm long, and `FALSE` values indicate rows where it is not (in `R`, "not equal to" is denoted as `!=`).

Finally, once we've found these rows, we examine all the information related to those alligators. When using square brackets, remember that the row number comes before the comma, and the column number comes after (e.g. `ali[row, column]`). If we omit either the row or column number, `R` will return all rows or all columns respectively. For instance, to view all the information (i.e. all columns) for the 30th row, you would use:

```{r, results="hide"}
ali[30,]
```

To make life a bit easier for ourselves, we can use these square brackets in combination with the conditional statement to extract the entire row for the alligator that's `r max(ali$length)` cm long. The `TRUE` and `FALSE` values act as indicators for which rows we want to display: if a value is `TRUE`, we see that row; if it's `FALSE`, we don't. We can achieve this by writing:

```{r, results="hide"}
ali[ali$length == 2293,]
```

What do you make of this observation? We're expecting `length` to be about 200 to 400 centimeters (2 to 4 m), but this alligator is apparently 10 times bigger than that. And `bite` is about 100 times bigger. Neither of the two values seems to have a decimal point (or at least `R` has given them zeros after the decimal point).

Why might that be the case? Let's find a "normal" observation to compare this giant alligator to. Assuming the first observation is a normal one, use the square brackets to extract the first row of data from the `ali` dataset:

```{r, results="hide"}
ali[1, ]
```

Well, they're clearly different. The first alligator in our dataset looks pretty normal. Maybe a bit on the big side (just over 4 m long), but not weirdly so. There are only two possible explanations for this: either there are giant alligators among us, or there has been a data entry mistake.

## Question Set Three

<div class="question-block">
<p><strong>Question 1.</strong> What may have occurred to cause this apparent data entry mistake?</p>
<div class="mc-options">
  <label><input type="radio" name="q3_q1"> The data was deliberately falsified</label>
  <label><input type="radio" name="q3_q1" data-correct="true"> Decimal points were accidentally omitted during data entry</label>
  <label><input type="radio" name="q3_q1"> The measurements were recorded in the wrong units</label>
  <label><input type="radio" name="q3_q1"> A duplicate row was accidentally added to the dataset</label>
</div>
<button class="reveal-btn" onclick="revealAnswer(this)">Show Answer</button>
<div class="instructor-answer">
<p>The decimal points were accidentally omitted during data entry. The correct values are 229.3 cm and 1236.32 PSI. If you look at the raw data file, neither value has a decimal point.</p>
</div>
</div>

# A More Refined Detective

The approach above works, but typing 2293 every time is awkward and can cause issues. Referencing specific numbers like this in code is called "magic numbers" because to anyone looking at your code, they have absolutely no idea why this number is special and what why you are working with it. Better to store the maximum in a clearly labelled variable and reference that instead:

```{r, results="hide"}
# Get the maximum alligator length
max_length <- max(ali$length)

# Now use this value to find the problematic rows
ali[ali$length == max_length, ]
```

By storing the result of `max(ali$length)` in a variable (`max_length`), we can simplify our code and make it more readable. This way, if we need to refer to the maximum length multiple times, we only need to use the `max_length` variable. Within the `ali` dataset, this code extracts all columns (since we have not specified anything after the comma in the square brackets) for any rows where the length of an alligator is equal to the maximum length of any alligator.

We can also adapt the conditional statement to find rows based on different criteria. For example, if we wanted to extract all rows with alligators that were greater than the average length, we could do:

```{r, results="hide"}
# Calculate the average alligator length
average_length <- mean(ali$length)

# Extract all rows where the alligator length is greater than the average
ali[ali$length > average_length, ]
```

In this case, `ali[ali$length > average_length, ]` will give us all columns for rows where the length of the alligator is greater than the calculated average length. We now get data from more than one alligator, as more than one alligator meets our condition of being longer than average. We can also see the "data entry mistake alligator" using this method.

Let's refine our approach further by using a different function: `quantile()`. First, rerun the `summary(ali)` code to observe the quantiles (`1st Qu.` and `3rd Qu.`) that it reports. These quantiles represent the values below which 25% of the data falls and above which 75% of the data falls. We can retrieve these quantile values ourselves by doing:

```{r, results="hide"}
quantile(ali$length, probs = c(0.25,0.75))
```

The `probs` argument in the `quantile()` function allows us to specify which probabilities we want to calculate quantiles for. In the previous example, we requested the 25% (`0.25`) and 75% (`0.75`) quantiles. Since we wanted to get these two values, we used the `c()` function, which combines (or "concatenates") multiple values into a single vector. If we only need a single quantile, such as the 75th percentile, we can directly write `quantile(ali$length, probs = 0.75)`. In this case, there's no need to use `c()` because we're only asking for one value.

## Question Set Four

<div class="question-block">
<p><strong>Question 1.</strong> Write the code that would extract all rows in the <code>ali</code> dataset where <code>bite</code> values are above the 3rd quantile.</p>
<textarea class="student-answer" rows="3" placeholder="Write your answer here..."></textarea>
<button class="reveal-btn" onclick="revealAnswer(this)">Show Answer</button>
<div class="instructor-answer">
<pre><code class="language-r">ali[ali&#36;bite > quantile(ali&#36;bite, probs = 0.75),]</code></pre>
</div>
</div>

<div class="question-block">
<p><strong>Question 2.</strong> Repeat the above, but for <code>bite</code> values <em>less than or equal to</em> the 33% quantile. If you're unsure how to specify "less than or equal to", have a quick search online or use your preferred LLM.</p>
<textarea class="student-answer" rows="3" placeholder="Write your answer here..."></textarea>
<button class="reveal-btn" onclick="revealAnswer(this)">Show Answer</button>
<div class="instructor-answer">
<pre><code class="language-r">ali[ali&#36;bite <= quantile(ali&#36;bite, probs = 0.33),]</code></pre>
</div>
</div>

<div class="question-block">
<p><strong>Question 3.</strong> Using the tools from above, can you identify any other possible data entry mistakes caused by missing decimal points?</p>
<div class="mc-options">
  <label><input type="radio" name="q4_q3"> Yes, there's at least one more observation with missing decimal points</label>
  <label><input type="radio" name="q4_q3" data-correct="true"> No, row 20 appears to be the only case</label>
</div>
<button class="reveal-btn" onclick="revealAnswer(this)">Show Answer</button>
<div class="instructor-answer">
<p>Row 20 appears to be the only observation with this problem. The other bite and length values all look plausible.</p>
</div>
</div>

# Correcting Data

So, we've identified a problem. What now? Do we delete the observation? Declare that [*Sarcosuchus*](https://en.wikipedia.org/wiki/Sarcosuchus) are not extinct and are alive and well in Parker Island Gator Farm, Florida, USA (and have actually doubled in size)? Set up a Facebook group to share our concrete and irrefutable evidence of *Sarcosuchus* and start producing a [film](https://www.facebook.com/cryptofilm/) to share this wonderful news with the world? Obviously not. That'd be silly.

But we do need to do something. Row 20 of our dataset is clearly a mistake, but correcting it is a really sensitive topic. Part of the reason for that sensitivity is because a non-trivial number of scientists have been found to actually *fake* their data in order to further their own careers and prestige (e.g. see [here](https://www.science.org/content/article/university-investigation-found-prominent-spider-biologist-fabricated-falsified-data), [here](https://www.science.org/content/article/does-ocean-acidification-alter-fish-behavior-fraud-allegations-create-sea-doubt), and most ironically [here](https://behavioralscientist.org/harvard-professor-under-scrutiny-for-alleged-data-fraud/) for examples).

We want to be better than that. So before touching anything, go back to the source: not the downloaded file, but the original recorded values. Here that means contacting Parker Island Gator Farm to confirm our suspicion that row 20 is simply missing decimal points. (Please don't actually contact them. They'll be confused as hell.)

In the interest of time, I'll confirm that it's missing decimal points so we can move on to fixing the problem. Specifically, `bite` should have two decimal points, and `length` should have one.

We have two options: edit `alligator_bite.txt` directly and reload, or fix it in `R`. Do it in `R`. If you edit the raw file, there's no record of what changed or why, and that starts to look a lot like the data manipulation in the examples I showed above.

Doing the correction in your code is totally transparent; it's right there for anyone who looks at your code to see.

So how do we fix row 20? Well, we use the skills from above: extracting, indexing, assigning, and simple math.

Let's start with `ali$bite`. We only want to change row 20. How do we select only row 20? `ali$bite[20]`. Note that we don't need to specify a column in the square brackets, because we're working with a single column, the `bite` column.

Next, we need to replace `ali$bite[20]` with an altered version of itself, so we would write:

```{r, results="hide"}
ali$bite[20] <- ali$bite[20]
```

This will just replace the `bite` value from row 20 with itself. We're not actually doing anything here, but we're getting closer.

Next, we need to "add" back in the decimal point that was lost during data entry. How do we shift the value two places to the right of the decimal point? We divide by 100, leaving us with:

```{r, results="hide"}
# Resolving data entry issue with missing decimal point
ali$bite[20] <- ali$bite[20] / 100
```

Using a hardcoded row number like `[20]` works here, but it's fragile: if the dataset were ever re-sorted or a row added above this one, row 20 would point to the wrong observation (i.e. 20 is a "magic number"). A safer approach uses the same conditional indexing we've already learned, targeting the value itself rather than its position:

```{r, eval=FALSE}
ali$bite[ali$bite > 100000] <- ali$bite[ali$bite > 100000] / 100
```

This reads as: "for any `bite` value greater than 100,000, divide it by 100." It will find the right row regardless of where it sits in the dataset.

## Question Set Five

<div class="question-block">
<p><strong>Question 1.</strong> Write (copy &amp; paste) the code you used to resolve the data entry mistake for <code>ali&#36;length</code>. Be sure to add a clear comment indicating that you have made this change, so that if anyone reviews your code, they understand the reason for the modification and don't suspect any foul play.</p>
<textarea class="student-answer" rows="3" placeholder="Write your answer here..."></textarea>
<button class="reveal-btn" onclick="revealAnswer(this)">Show Answer</button>
<div class="instructor-answer">
<pre><code class="language-r"># Correcting missing decimal point, confirmed by Parker Island Gator Farm on 1/1/2020
ali&#36;length[ali&#36;length > 2000] <- ali&#36;length[ali&#36;length > 2000] / 10</code></pre>
</div>
</div>

<div class="question-block">
<p><strong>Question 2.</strong> Having made any changes to your data, you need to check it to make sure you didn't create a new mistake. Do so now using any techniques you feel are appropriate. Copy and paste the code you used to check if the correction worked.</p>
<textarea class="student-answer" rows="3" placeholder="Write your answer here..."></textarea>
<button class="reveal-btn" onclick="revealAnswer(this)">Show Answer</button>
<div class="instructor-answer">
<pre><code class="language-r">summary(ali)</code></pre>
</div>
</div>

# Do it for Real

Download the data file below and save it to your `BI3010/data/` folder.

```{r, echo=FALSE, results='asis'}
cat(make_download_btn("data/animal_speed.txt"))
```

The goal is the same as before: find all the data entry mistakes. This dataset is larger and the errors are harder to spot, so you'll need to be more systematic. A few hints:

1. `unique()` lists every distinct value in a column, useful for spotting typos in text columns.
2. `mean()` returns `NA` if any values are missing. Add `na.rm = TRUE` to ignore them: `mean(variable, na.rm = TRUE)`.
3. There are five mistakes in total.

Start with an EDA as before. Can you find them all?

::: {.callout-note}
## Information about the dataset

| Variable | Description |
|---|---|
| `species` | One of three species: Peregrine Falcon, Saiga Antelope, or Cheetah |
| `weight` | Body weight in grams |
| `speed` | Speed over a 50 m distance, in km/h |

- A total of 500 individuals were measured, but these weren't equally divided between the three species.
- Multiple people participated in data entry, and it's believed that not all were confident in how to do this.
- Physical copies of the data are available should any data need to be verified (contact one of the teaching staff if you need a data entry confirmed).
:::

```{r, eval=FALSE}
fast <- read.table("data/animal_speed.txt", header = TRUE)
```

```{r, results="hide"}
# Start your exploration here
summary(fast)
str(fast)
unique(fast$species)
```

<div class="question-block">
<p><strong>Question 1.</strong> Identify all unusual observations in the dataset.</p>
<textarea class="student-answer" rows="6" placeholder="List every unusual observation you found..."></textarea>
<button class="reveal-btn" onclick="revealAnswer(this)">Show Answer</button>
<div class="instructor-answer">
<ul>
<li>One individual has a recorded weight of 0 g. It should be 45514.</li>
<li>One individual has a recorded weight of 3 g. It should be 47255.</li>
<li>One of the species names has been entered incorrectly. It should be Peregrine Falcon.</li>
<li>One Peregrine Falcon has a recorded speed of 35,711 km/h. It is missing two decimal places.</li>
<li>One observation has a missing value (<code>NA</code>, or "not available") in the speed column. The data provider does not know what this value should be.</li>
</ul>
</div>
</div>

<div class="question-block">
<p><strong>Question 2.</strong> Copy and paste the code you used to resolve the above issues.</p>
<textarea class="student-answer" rows="6" placeholder="Write your correction code here..."></textarea>
<button class="reveal-btn" onclick="revealAnswer(this)">Show Answer</button>
<div class="instructor-answer">
<pre><code class="language-r"># Weight of 0 g corrected to the value confirmed by the data provider
fast&#36;weight[fast&#36;weight == 0] <- 45514

# Weight of 3 g corrected to the value confirmed by the data provider
fast&#36;weight[fast&#36;weight == 3] <- 47255

# Species typo corrected, then re-factored
fast&#36;species[fast&#36;species == "Peregrin Took"] <- "Peregrine Falcon"
fast&#36;species <- factor(fast&#36;species)

# Peregrine Falcon speed missing two decimal places, so divide by 100
fast&#36;speed[fast&#36;speed > 10000 &amp; !is.na(fast&#36;speed)] <- fast&#36;speed[fast&#36;speed > 10000 &amp; !is.na(fast&#36;speed)] / 100

# The NA in speed is left as-is: the data provider cannot confirm the true value

# Final check
summary(fast)</code></pre>
</div>
</div>

## Question Set Six

::: {.callout-warning}
Make sure you have corrected all five data entry errors before answering these questions - the answers depend on the corrected dataset.
:::

<div class="question-block">
<p><strong>Question 1.</strong> What is the median weight in this dataset?</p>
<textarea class="student-answer" rows="2" placeholder="Write your answer here..."></textarea>
<button class="reveal-btn" onclick="revealAnswer(this)">Show Answer</button>
<div class="instructor-answer">
<p>The median weight is <strong>46,453 g</strong>.</p>
<pre><code class="language-r">median(fast&#36;weight)</code></pre>
</div>
</div>

<div class="question-block">
<p><strong>Question 2.</strong> What is the median speed in this dataset?</p>
<textarea class="student-answer" rows="2" placeholder="Write your answer here..."></textarea>
<button class="reveal-btn" onclick="revealAnswer(this)">Show Answer</button>
<div class="instructor-answer">
<p>The median speed is <strong>102.1 km/h</strong>.</p>
<pre><code class="language-r">median(fast&#36;speed, na.rm = TRUE)</code></pre>
</div>
</div>

<div class="question-block">
<p><strong>Question 3.</strong> Which species is most common in this dataset?</p>
<textarea class="student-answer" rows="2" placeholder="Write your answer here..."></textarea>
<button class="reveal-btn" onclick="revealAnswer(this)">Show Answer</button>
<div class="instructor-answer">
<p>The most common species is the <strong>Saiga Antelope</strong> (177 individuals, compared with 167 Cheetahs and 156 Peregrine Falcons). The <code>table()</code> function gives a count per species, but there are other ways to check:</p>
<pre><code class="language-r">table(fast&#36;species)</code></pre>
</div>
</div>

# Chupacabras, Aliens and Beer

The final dataset is `chupa.txt`. We'll work through it together in the last 30 minutes, but start early if you're ahead. It contains the following variables:

- `year`: The year in which data was collected
- `city`: The city where the data was collected
- `alc`: The average alcohol consumption in that city, in that year
- `ufo`: The number of [unidentified flying objects](https://en.wikipedia.org/wiki/Unidentified_flying_object) (i.e. aliens) reported in that city, in that year
- `belief`: The most common response from a selection of members of the public when asked if they believe in the supernatural, in that city, in that year
- `chupa`: The growth rate of [chupacabra](https://en.wikipedia.org/wiki/Chupacabra) sightings from the previous year to the current year, in that city (where a value of zero means the number of sightings is constant, -0.5 means the number of sightings has declined by 50%, and 0.5 means they have increased by 50%).

There are data entry errors in this dataset. Find them all using EDA.

Download the data file below and save it to your `BI3010/data/` folder.

```{r, echo=FALSE, results='asis'}
cat(make_download_btn("data/chupa.txt"))
```

As a group, we'll look at snippets of code and identify the errors.

```{r, eval=FALSE}
chupa <- read.table("data/chupa.txt", header = TRUE)
```

```{r, results="hide"}
head(chupa)
str(chupa)
unique(chupa$city)
summary(chupa)
unique(chupa$belief)
```

<div class="question-block">
<p><strong>Find and fix all data entry errors</strong> in the <code>chupa</code> dataset. Describe each error and write the correction code.</p>
<textarea class="student-answer" rows="6" placeholder="Describe each error and your correction code here..."></textarea>
<button class="reveal-btn" onclick="revealAnswer(this)">Show Answer</button>
<div class="instructor-answer">
<p><strong>Error 1 - Two spellings of Amarillo</strong><br>
One row has the city entered as "amarillo" (lower case a) while all other rows use "Amarillo". Use <code>unique(chupa$city)</code> to spot this, then check the rows affected with indexing.</p>

<p><strong>Error 2 - Placeholder value in ufo (9999)</strong><br>
The <code>ufo</code> column contains one value of 9999, which appears to be a placeholder for missing data rather than a real count. A histogram of <code>ufo</code> makes this obvious. This value should be replaced with <code>NA</code>.</p>

<p><strong>Error 3 - Negative value in alc</strong><br>
Alcohol consumption can't be negative. There's one observation with a value of -5 in the <code>alc</code> column. A histogram will reveal it as a clear outlier on the left. The correct value is 5. Note: negative values in the <code>chupa</code> column are fine and expected, since it's a growth rate.</p>
</div>
</div>

# Coding Puzzles

In the following snippets of code, can you identify the error? If you're not sure, try running them and see what the error says.

<div class="puzzle-block">
<p><strong>Puzzle 1.</strong></p>
<pre><code class="language-r">summary(alc)</code></pre>
<textarea class="student-answer" rows="2" placeholder="What is the error?"></textarea>
<button class="reveal-btn" onclick="revealAnswer(this)">Show Answer</button>
<div class="puzzle-answer">
<pre><code class="language-r"># We have not specified that this is in the chupa dataset
summary(chupa&#36;alc)</code></pre>
</div>
</div>

<div class="puzzle-block">
<p><strong>Puzzle 2.</strong></p>
<pre><code class="language-r">chupa[chupa&#36;ufo==9999]</code></pre>
<textarea class="student-answer" rows="2" placeholder="What is the error?"></textarea>
<button class="reveal-btn" onclick="revealAnswer(this)">Show Answer</button>
<div class="puzzle-answer">
<pre><code class="language-r"># We're missing the comma to specify which columns to show
chupa[chupa&#36;ufo == 9999, ]</code></pre>
</div>
</div>

<div class="puzzle-block">
<p><strong>Puzzle 3.</strong></p>
<pre><code class="language-r">chupa[chupa&#36;ufo = 9999, ]</code></pre>
<textarea class="student-answer" rows="2" placeholder="What is the error?"></textarea>
<button class="reveal-btn" onclick="revealAnswer(this)">Show Answer</button>
<div class="puzzle-answer">
<pre><code class="language-r"># = assigns a value; use == for a conditional test
chupa[chupa&#36;ufo == 9999, ]</code></pre>
</div>
</div>

<div class="puzzle-block">
<p><strong>Puzzle 4.</strong></p>
<pre><code class="language-r">chupa[chupa&#36;city = Houston,]</code></pre>
<textarea class="student-answer" rows="2" placeholder="What is the error?"></textarea>
<button class="reveal-btn" onclick="revealAnswer(this)">Show Answer</button>
<div class="puzzle-answer">
<pre><code class="language-r"># Houston needs to be in quotation marks (it's a string, not an object)
# Also = should be ==
chupa[chupa&#36;city == "Houston", ]</code></pre>
</div>
</div>

<div class="puzzle-block">
<p><strong>Puzzle 5.</strong></p>
<pre><code class="language-r">ali[ali&#36;Length > 200, ]</code></pre>
<textarea class="student-answer" rows="2" placeholder="What is the error?"></textarea>
<button class="reveal-btn" onclick="revealAnswer(this)">Show Answer</button>
<div class="puzzle-answer">
<pre><code class="language-r"># Names are case sensitive. Should be `length`, not `Length`
ali[ali&#36;length > 200, ]</code></pre>
</div>
</div>

<div class="puzzle-block">
<p><strong>Puzzle 6.</strong></p>
<pre><code class="language-r">ali[ali&#36;bite > mean(ali&#36;bite), ]</code></pre>
<textarea class="student-answer" rows="2" placeholder="What is the error?"></textarea>
<button class="reveal-btn" onclick="revealAnswer(this)">Show Answer</button>
<div class="puzzle-answer">
<pre><code class="language-r"># If there are NAs in bite this will fail - include na.rm = TRUE
ali[ali&#36;bite > mean(ali&#36;bite, na.rm = TRUE), ]</code></pre>
</div>
</div>

```{=html}
<button id="glossary-btn" onclick="toggleGlossary()">Glossary / 词汇表</button>

<div id="glossary-panel">
  <div id="glossary-header">
    <span>Glossary / 词汇表</span>
    <button id="glossary-close" onclick="toggleGlossary()">&#x2715;</button>
  </div>
  <div id="glossary-body">

    <div class="glossary-section-header">Workflow</div>

    <div class="glossary-entry">
      <div class="glossary-terms">
        <span class="glossary-en-term">Exploratory Data Analysis (EDA)</span>
        <span class="glossary-zh-term">探索性数据分析</span>
      </div>
      <div class="glossary-en-def">An initial investigation of a dataset using summaries and plots to understand its structure, spot patterns, and identify potential problems before modelling.</div>
      <div class="glossary-zh-def">在建模之前，通过统计摘要和图表初步了解数据结构、发现规律并识别潜在问题的过程。</div>
    </div>

    <div class="glossary-entry">
      <div class="glossary-terms">
        <span class="glossary-en-term">Data entry error</span>
        <span class="glossary-zh-term">数据录入错误</span>
      </div>
      <div class="glossary-en-def">A mistake introduced when recording or entering data (e.g. a typo, wrong units, a placeholder value left in). EDA is the main tool for catching these.</div>
      <div class="glossary-zh-def">在记录或输入数据时引入的错误（如错别字、单位错误、遗留的占位符值）。探索性数据分析是发现此类错误的主要方法。</div>
    </div>

    <div class="glossary-section-header">Data structure</div>

    <div class="glossary-entry">
      <div class="glossary-terms">
        <span class="glossary-en-term">Data frame</span>
        <span class="glossary-zh-term">数据框</span>
      </div>
      <div class="glossary-en-def">R's standard table structure: rows are observations, columns are variables. Most datasets you work with in R will be data frames.</div>
      <div class="glossary-zh-def">R 中最常用的表格结构：行为观测值，列为变量。大多数在 R 中处理的数据集都是数据框。</div>
    </div>

    <div class="glossary-entry">
      <div class="glossary-terms">
        <span class="glossary-en-term">Observation</span>
        <span class="glossary-zh-term">观测值</span>
      </div>
      <div class="glossary-en-def">A single row in a dataset, representing one instance of data collection (e.g. one animal, one survey response).</div>
      <div class="glossary-zh-def">数据集中的一行，代表一次数据采集的记录（如一只动物、一份问卷回答）。</div>
    </div>

    <div class="glossary-entry">
      <div class="glossary-terms">
        <span class="glossary-en-term">Variable</span>
        <span class="glossary-zh-term">变量</span>
      </div>
      <div class="glossary-en-def">A column in a dataset. Each variable holds one type of measurement or attribute recorded for every observation.</div>
      <div class="glossary-zh-def">数据集中的一列，记录每个观测值的某种测量或属性。</div>
    </div>

    <div class="glossary-section-header">Data types</div>

    <div class="glossary-entry">
      <div class="glossary-terms">
        <span class="glossary-en-term">Numeric</span>
        <span class="glossary-zh-term">数值型</span>
      </div>
      <div class="glossary-en-def">A variable type that holds numbers (e.g. length, weight, temperature). R calls these <code>num</code> or <code>int</code> in <code>str()</code> output.</div>
      <div class="glossary-zh-def">存储数字的变量类型（如长度、重量、温度）。在 R 的 <code>str()</code> 输出中显示为 <code>num</code> 或 <code>int</code>。</div>
    </div>

    <div class="glossary-entry">
      <div class="glossary-terms">
        <span class="glossary-en-term">Character</span>
        <span class="glossary-zh-term">字符型</span>
      </div>
      <div class="glossary-en-def">A variable type that holds text (e.g. species names, city names). R calls these <code>chr</code> in <code>str()</code> output.</div>
      <div class="glossary-zh-def">存储文本的变量类型（如物种名称、城市名称）。在 R 的 <code>str()</code> 输出中显示为 <code>chr</code>。</div>
    </div>

    <div class="glossary-entry">
      <div class="glossary-terms">
        <span class="glossary-en-term">Factor</span>
        <span class="glossary-zh-term">因子</span>
      </div>
      <div class="glossary-en-def">A categorical variable with a fixed set of levels (e.g. sex: male/female). R stores factors efficiently and uses them correctly in models.</div>
      <div class="glossary-zh-def">具有固定类别水平的分类变量（如性别：雄性/雌性）。R 以因子形式高效存储并在模型中正确使用。</div>
    </div>

    <div class="glossary-section-header">Summary statistics</div>

    <div class="glossary-entry">
      <div class="glossary-terms">
        <span class="glossary-en-term">Mean</span>
        <span class="glossary-zh-term">均值</span>
      </div>
      <div class="glossary-en-def">The arithmetic average: sum of all values divided by the number of values. Sensitive to extreme outliers.</div>
      <div class="glossary-zh-def">算术平均数：所有值的总和除以值的个数。对极端离群值较为敏感。</div>
    </div>

    <div class="glossary-entry">
      <div class="glossary-terms">
        <span class="glossary-en-term">Median</span>
        <span class="glossary-zh-term">中位数</span>
      </div>
      <div class="glossary-en-def">The middle value when data are sorted. More robust than the mean when outliers are present.</div>
      <div class="glossary-zh-def">数据排序后的中间值。在存在离群值时比均值更稳健。</div>
    </div>

    <div class="glossary-entry">
      <div class="glossary-terms">
        <span class="glossary-en-term">Quantile</span>
        <span class="glossary-zh-term">分位数</span>
      </div>
      <div class="glossary-en-def">A cut point that divides data into equal-sized groups. The 25th and 75th percentiles (Q1 and Q3) bracket the middle half of the data.</div>
      <div class="glossary-zh-def">将数据分成等份的分界点。第25和第75百分位数（Q1和Q3）框定了数据的中间一半。</div>
    </div>

    <div class="glossary-entry">
      <div class="glossary-terms">
        <span class="glossary-en-term">Range</span>
        <span class="glossary-zh-term">极差</span>
      </div>
      <div class="glossary-en-def">The difference between the maximum and minimum values in a variable. A quick check for implausible values.</div>
      <div class="glossary-zh-def">变量中最大值与最小值之差。可快速检查是否存在不合理的值。</div>
    </div>

    <div class="glossary-section-header">Missing and unusual values</div>

    <div class="glossary-entry">
      <div class="glossary-terms">
        <span class="glossary-en-term">NA</span>
        <span class="glossary-zh-term">缺失值</span>
      </div>
      <div class="glossary-en-def">R's symbol for a missing value. Functions like <code>mean()</code> return <code>NA</code> if any value is missing unless you add <code>na.rm = TRUE</code>.</div>
      <div class="glossary-zh-def">R 中缺失值的符号。若数据中存在 <code>NA</code>，<code>mean()</code> 等函数会返回 <code>NA</code>，除非添加 <code>na.rm = TRUE</code>。</div>
    </div>

    <div class="glossary-entry">
      <div class="glossary-terms">
        <span class="glossary-en-term">Outlier</span>
        <span class="glossary-zh-term">离群值</span>
      </div>
      <div class="glossary-en-def">A value that sits far from the rest of the data. May be a genuine extreme observation or a data entry error.</div>
      <div class="glossary-zh-def">远离其他数据点的值。可能是真实的极端观测值，也可能是数据录入错误。</div>
    </div>

    <div class="glossary-section-header">R programming</div>

    <div class="glossary-entry">
      <div class="glossary-terms">
        <span class="glossary-en-term">Script</span>
        <span class="glossary-zh-term">脚本</span>
      </div>
      <div class="glossary-en-def">A plain text file (usually <code>.R</code>) containing a sequence of R commands. Run it top to bottom to reproduce an analysis exactly.</div>
      <div class="glossary-zh-def">包含一系列 R 命令的纯文本文件（通常为 <code>.R</code>）。从上到下运行即可完整重现分析过程。</div>
    </div>

    <div class="glossary-entry">
      <div class="glossary-terms">
        <span class="glossary-en-term">Function</span>
        <span class="glossary-zh-term">函数</span>
      </div>
      <div class="glossary-en-def">A named operation in R (e.g. <code>mean()</code>, <code>summary()</code>). You pass inputs inside the parentheses and get an output back.</div>
      <div class="glossary-zh-def">R 中的命名操作（如 <code>mean()</code>、<code>summary()</code>）。在括号内传入输入值，函数返回输出结果。</div>
    </div>

    <div class="glossary-entry">
      <div class="glossary-terms">
        <span class="glossary-en-term">Argument</span>
        <span class="glossary-zh-term">参数</span>
      </div>
      <div class="glossary-en-def">An input you provide to a function, e.g. <code>na.rm = TRUE</code> in <code>mean(x, na.rm = TRUE)</code>. Arguments control how the function behaves.</div>
      <div class="glossary-zh-def">传递给函数的输入值，例如 <code>mean(x, na.rm = TRUE)</code> 中的 <code>na.rm = TRUE</code>。参数控制函数的行为方式。</div>
    </div>

    <div class="glossary-entry">
      <div class="glossary-terms">
        <span class="glossary-en-term">Indexing</span>
        <span class="glossary-zh-term">索引</span>
      </div>
      <div class="glossary-en-def">Selecting specific rows or columns from a data frame using square brackets: <code>df[rows, columns]</code>. Leave a slot blank to mean "all".</div>
      <div class="glossary-zh-def">使用方括号从数据框中选取特定的行或列：<code>df[行, 列]</code>。留空表示"全部"。</div>
    </div>

    <div class="glossary-entry">
      <div class="glossary-terms">
        <span class="glossary-en-term">Conditional statement</span>
        <span class="glossary-zh-term">条件语句</span>
      </div>
      <div class="glossary-en-def">A logical test that returns <code>TRUE</code> or <code>FALSE</code> (e.g. <code>x > 5</code>). Used inside square brackets to filter rows that meet a condition.</div>
      <div class="glossary-zh-def">返回 <code>TRUE</code> 或 <code>FALSE</code> 的逻辑判断（如 <code>x > 5</code>）。用于方括号内筛选满足条件的行。</div>
    </div>

    <div class="glossary-entry">
      <div class="glossary-terms">
        <span class="glossary-en-term">Working directory</span>
        <span class="glossary-zh-term">工作目录</span>
      </div>
      <div class="glossary-en-def">The folder R looks in by default when reading or writing files. Set it with <code>setwd()</code> or via the RStudio menu.</div>
      <div class="glossary-zh-def">R 读写文件时默认查找的文件夹。可通过 <code>setwd()</code> 或 RStudio 菜单设置。</div>
    </div>

  </div>
</div>
```
